---
title: "Introduction to Lavaan"
author: "Amanda Montoya"
date: "06/17/2026"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is lavaan?

`lavaan` (latent variable analysis) is an R package developed by Yves Rosseel


It was first released in 2012


When you use `lavaan` you should cite it!


Yves Rosseel (2012). lavaan: An R Package for Structural Equation Modeling. Journal of Statistical Software, 48(2), 1-36. URL http://www.jstatsoft.org/v48/i02/



## Installing lavaan

`lavaan` is a package in R which is downloadable via CRAN


```{r, eval = FALSE}
install.packages("lavaan", dependencies = TRUE)
```

You only ever need to install `lavaan` once on each computer you use it on. But you should regularly update your packages. Always report the version number you are using in your paper!


If `lavaan` is correctly installed you should get a little happy message when you run the following:


```{r}
library(lavaan)
```


You do need to run the above command every time you start an R session. 


I recommend in your scripts always having a section at the top which includes all your libraries. 


## Model Specification


`lavaan` works by creating an object in R which is a string (characters) that takes on a specific form to specify your model. 


That string is then input into one of the lavaan functions (more on this later) along with the data to fit the model. 


Here's an example:


```{r, eval = FALSE}
myModel <- '
   # latent variables
     ind60 =~ x1 + x2 + x3
     dem60 =~ y1 + y2 + y3 + y4
     dem65 =~ y5 + y6 + y7 + y8
   # regressions
     dem60 ~ ind60
     dem65 ~ ind60 + dem60
   # residual covariances
     y1 ~~ y5
     y2 ~~ y4 + y6
     y3 ~~ y7
     y4 ~~ y8
     y6 ~~ y8
'
fit <- sem(model = myModel,
           data  = PoliticalDemocracy)
summary(fit)
```


We will start by learning about the different symbols and how they are used in the syntax. 


### Regression ($B$)


To specify the paths that are included in $B$ you will need to use the syntax ~ ("tilde" pronounced "til-duh")


Variables on the left side of the ~ are the endogenous variables. 


Variable on the right side of the ~ are the predictors (could be endogenous or exogenous).


```{r, eval = FALSE}
y1~x1
```


This command means the variable `y1` in the specified data set has one predictor: `x1`


You can have multiple predictors by putting a `+` in between their names


```{r, eval = FALSE}
y1~x1+x2
```


This command means the variable `y1` in the specified data set has two predictors: `x1` and `x2`


You can have multiple regression equations


```{r, eval = FALSE}
y1~x1+x2
y2~x1+x2+x3
```


Now we have two endogenous variables: `y1` and `y2`. 


Here the equation for `y1` is the same as the previous example.


Additionally `y2` has 3 predictors `x1`, `x2`, and `x3`. 


The predictors for each endogenous variable can be the same or different from the other equations. 


If you have two endogenous variables which have the same equations you can shorten the syntax


```{r, eval = FALSE}
y1~x1+x2+x3
y2~x1+x2+x3
```


is the same as


```{r, eval = FALSE}
y1+y2~x1+x2+x3
```


be very careful with this though. 


Remember that endogenous variables can also be predictors, so we could have:


```{r, eval = FALSE}
y1~x1
y2~y1+x1
```


The variables in these equations can be latent variables as well, but we'll leave that for a future lesson. 


## Variances and Covariances ($\Sigma_\epsilon\epsilon$)


You can specify that two variables covary by using `~~`


For example if I want to specify that `x1` and `x2` covary I will specify


```{r, eval=FALSE}
x1~~x2
```


If I want to estimate the variance of `x1` (which is the covariance of a variable with itself)


```{r, eval=FALSE}
x1~~x1
```


*Important Warning: Covariances of deviations*


Perhaps one of the least intuitive things about `lavaan` is the way we specify covariances of deviances. 


Consider the following model:


```{r, eval = FALSE}
y1~x1+x2
y2~x1+x2+x3
```


Let's say I want to let the deviance for `y1` covary with the deviance for `y2`. 


I could add


```{r, eval=FALSE}
y1~~y2
```


This would add a covariance to the residuals (not the variables directly). 


If you want to incorporate multiple covariances you can express it two ways


```{r, eval=FALSE}
y1~~y2
y1~~y3
```


or 


```{r, eval=FALSE}
y1~~y2+y3
```


## Putting the whole script together


To specify the whole command you will put each line together in a string. 


You can use line breaks or semicolons to separate the commands


```{r, eval=FALSE}
modelsyntax <- '
     y1~x1+x2
     y2~x1+x2+x3
     y1~~y2
'
```


or 


```{r, eval=FALSE}
modelsyntax <- 'y1~x1+x2; y2~x1+x2+x3; y1~~y2'
```


You will likely want to save the string as an object (`name <- `)


You can also save different parts separately (this is useful for fitting multiple models with different parts)


```{r, eval=FALSE}
regressions <- "
     y1~x1+x2
     y2~x1+x2+x3
"
covariances <- "y1~~y2"
```


You can include comments inside the script using `#` to help remind yourself what everything is and stay organized!


```{r, eval=FALSE}
modelsyntax <- "
     #regressions
     y1~x1+x2
     y2~x1+x2+x3
     
     #covariances
     y1~~y2
"
```


## Fitting Functions in `lavaan`


`lavaan` has multiple functions built in to help with estimating models. The three we will use in this class are 

* lavaan
* sem
* cfa


The `lavaan()` function is the most general, it essentially requires full specification (nothing is done for you)


The `sem()` function is probably what you will use most. It allows for latent variable regressions, whereas `cfa()` does not. The `cfa()` function is specifically for latent variable models with no regressions among the latent variables. 


`sem()` and `cfa()` are considered "helper functions", this means they automatically do a few things for us without us specifying them:

* Fix the latent variable intercepts to zero ($\alpha_L$)
* Elements of $\Sigma_{\epsilon_{L_X} \epsilon_{L_X}}$ and $\Sigma_{\epsilon_{L_Y} \epsilon_{L_Y}}$ are freely estimated unless otherwise specified (unless this results in an underidentified)


## Introducing our data


The data we'll use today come from Beshai et al. (2020) (https://doi.org/10.1016/j.brat.2020.103724). The study looks at the effectiveness of a Mindfulness Intervention in reducing depression and anxiety, by increasing mindfulness and self-compassion. All measures were taken at pre-test and post-test.


For now we'll stick with 4 variables:


* PreMindfulness: Mindfulness at pretest as measured by a scale score on the Five Facet Mindfulness Questionnaire - 15
* PostMindfulness: Mindfulness at posttest as measured by a scale score on the Five Facet Mindfulness Questionnaire - 15
* PreAnxiety: Anxiety at pretest as measured by a scale score on the Generalized Anxiety Disorder - 7. 
* PostAnxiety: Anxiety at posttest as measured by a scale score on the Generalized Anxiety Disorder - 7. 


Let's load in the data


```{r}
BeshaiData <- read.csv("BeshaiData.csv", header = TRUE)
names(BeshaiData)[1] <- "PreMindfulness"
```


## Fitting the Model


First, let's consider a simple model, where PreMindfulness predicts PostMindfulness. 


Specify the model syntax


```{r}
simplesyntax <- '
    PostMindfulness~PreMindfulness
    '
```


The equation for this would be:


$$Z_2 = \alpha_2+B_{21}Z_1+\epsilon_2$$


Next we'll use a fitting function (`sem()`) to fit the model and save the result


```{r}
simplefit <- sem(simplesyntax, data = BeshaiData)
```


Now we'll look at the summary for `simplefit`


```{r}
summary(simplefit)
```


The estimator is ML means that we used maximum likelihood to estimate the model. This is the default in lavaan, but we could also specify the argument `estimator = "ML"` is the `sem()` command and get the same result. 


Number of model parameters is the number of parameters estimates ($N_\theta$)


Number of observations is the sample size ($N$)


Ignore "Model Test User Model" for now... we'll get back to this soon


The "Parameter Estimates" section describes what matrices were used for estimation. We'll stick with the defaults for now. 


Under regressions, we get a path coefficient, standard error, z-value, and p-value! This is $\hat{B_{21}}$, with it's estimated standard errors and significance test. 


In the variances, we get the estimated variance of $\epsilon_2$ (i.e., the residual variance of $Z_2$). You can tell it's a residual variance because of the . infront of the variable name (.PostMindfulness)


## Extracting Information


### lavInspect


The `lavInspect()` function can be used to "look inside" a specific model object. It will print out the model matrices that are used for the model. 


Free parameters are non-zero integers. Fixed parameters are 0 (even if they are fixed to something other than zero)


```{r}
lavInspect(simplefit)
```


The notation that `lavaan` uses is kind of a modified LISREL notation (closer to the all-y LISREL notation)


### fitted.values


The `fitted.value()` function returns the model-implied covariance matrix


```{r}
fitted.values(simplefit)
```


### vcov


`vcov` gives the estimated covariance matrix of the *parameter estimates* (that second derivative thing we talked about last time). The square root of the diagonal of this matrix is the standard errors. 


```{r}
vcov(simplefit)
sqrt(diag(vcov(simplefit)))
```


## Model Constraints

### Fixing Parameters at specific values

You can fix certain parameters to specific values by specifying them in the model syntax. You do this by entering in the number and * as part of the specification.

```{r, eval = FALSE}
y1~~0*y2
```

This will set the deviances of y1 and y2 to be uncorrelated 

Doing this with deviances is very important because `lavaan` can be a little cavalier with correlated deviances. 

Similarly you can set specific regression coefficients to be a certain value. 

```{r, eval = FALSE}
y1~1*x1+x2
```

This will fix the regression coefficient of x1 to be 1, while the coefficient for x2 is freely estimated

### Fixing values to be equal

You can name specific parameters in `lavaan` with the same approach

```{r, eval = FALSE}
y1~B31*x1+B32*x2
```

And if you wanted to fix two parameters to be equal, give them the same name

```{r, eval = FALSE}
y1~B31*x1+B32*x2
y2~B31*x1+B32*x2
```

lavaan would only estimate two regression parameters in this model. 

## Practice with Beshai Data


Consider a model where PreMindfulness, PreAnxiety, and PostMindfulness all predict PostAnxiety. PreMindfulness predicts PostMindfulness.


Try to sketch out a path diagram before moving on. 


Write out the equations for PostMindfulness $y_1$, and PostAnxiety $y_2$


Model Syntax


```{r}
modelsyntax <- '
   # regressions
     PostMindfulness ~ PreMindfulness
     PostAnxiety ~ PreMindfulness+PreAnxiety+PostMindfulness
'
```


We have a model for PostMindfulness and a model for PostAnxiety.


```{r}
fit <- sem(model = modelsyntax,
           data  = BeshaiData)
summary(fit)
```

What did we learn?


* Higher levels of pretest mindfulness are significantly related to higher levels of posttest mindfulness ($B_{31} = 0.779$, $SE = 0.052$, $p < .001$)
* Higher levels of pretest mindfulness are significantly related to higher levels of posttest anxiety controlling for pretest anxiety and posttest mindfulness ($B_{41} = 0.167$, $SE = 0.055$, $p = .002$)
* Higher levels of pretest anxiety significantly related to higher levels of posttest anxiety controlling for pretest mindfulness and posttest mindfulness ($B_{42} = 0.558$, $SE = .065$, $p < .001$)
* Higher levels of posttest mindfulness are significantly related to lower levels of posttest anxiety controlling for pretest mindfulness and pretest mindfulness ($B_{43} = -0.338$, $SE = .052$, $p < .001$)


Variances of the deviances are significant, suggesting that there is residual variance (we have not explained all the variability in either outcome). *Note* this is usually a kind of trivial result that is not always of great interest. 


### Standardized Coefficients?


You can get **standardized coefficient estimates** by adding `standardized = TRUE` to the `summary` function. Two results will be added:

* Std.lv means that all latent variables are standardized
* Std.all means both latent and observed variables are standardized

```{r}
summary(fit, standardized = TRUE)
```


Notice that the `std.lv` results are no different from the original results, because there are not latent variables in the model. The `std.all` coefficients are different because the standard deviation of the observed variables is set to 1 (notice this also changes the variance estimates for the deviations). 


The deviation variances now have a nice interpretation


$1-\Sigma_{\epsilon_Y{_j} \epsilon_Y{_j}$ is the explained variance in $y_j$


59% of the variance in posttest mindfulness is explained by other variables in the model


49% of the variance in posttest anxiety is explained by other variables in the model


### Correlated Deviances?


In this model the deviances are not correlated. Let's see what happens if we allow them to be correlated. 


```{r}
corzeta <- 'PostMindfulness~~PostAnxiety'
fit2 <- sem(c(modelsyntax, corzeta), data = BeshaiData)
summary(fit2)
```


The model is not identified and `lavaan` has caught that! It provides some estimates (actually the same as if the correlation is fixed at zero) but no standard errors. 
